---
title: "Tutorial on niche-neutral GLMMs to understand community assembly"
author: "Sara Mortara et al."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    toc: true
    toc_float: true

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)#, cache=TRUE)
```

### The framework

The general idea of our framework is to translate niche and neutral
dynamics on community assembly into fixed and random effects of
generalized mixed models.  Within our framework one can simultaneously
test the importance of niche solely, neutrality solely and
combinations of niche and neutral effects on community structure
(abundance or occurrence).

We use the model selection approach to test multiple hypothesis and we
use conditional and marginal $R^{2}$ values to quantify the relative
importance of niche and neutrality on community structure.

Main criticism of our framework rely on: (1) Are fixed and random
effects actually capturing niche and neutral dynamics? (2) Are random
effects only capturing neutral dynamics or uninformed species traits
are inflating random effects?

In order to show how we implemented our framework and how model
components are translated into niche and neutral dynamics we performed
simulations as described below. We do not intend to scan all parameter
space of type of communities and traits, neither make a comprehensive
investigation of sampling effects. Our goal with this tutorial is to
address main criticisms of our framework and show how promising our
framework is to quantify the relative importance of niche and neutral
dynamics.

## A test of our modelling framework

Here, we present a working example of our modelling framework. We use
simulated communities based on niche and neutral dynamics in order to
show how fixed and random effects capture different ecological
processes. Also, we use traits with strong and weak correlations to
species abundance in order to show how fixed and random effects
capture neutral dynamics when traits are uninformative. Therefore, we
simulate stochastic and deterministic meta-communities and use strong
or weak traits in model selection.

We simulate meta-communities with the same data structure as our
abundance data of ferns in three mountain chains in southern
Brazil. Then, we make Poisson of communities and use samples from the
meta-communities in our model selection framework.

In order to address main criticisms of our framework, we analyze into three scenarios:

* Deterministic community with traits strongly correlated with species
  abundance from Poisson sample
* Deterministic community with traits poorly correlated with species
  abundance from Poisson sample
* Stochastic community with traits strongly correlated with species
  abundance from Poisson sample

<!-- Tiramos o: Stochastic community with right traits from Negative Binomial sample--> 


### 1. Building simulated communities

#### 1.1. Getting data for simulations

The simulations are based on several packages that should be installed
in the user's computer before starting the tutorial. The packages
bellow should be installed as regular CRAN packages before loaded. To
check if they are already installed run the code below.

```{r,  message=FALSE}
packages <- c("devtools", "ade4", "vegan", "lme4", "bbmle", "optimx",
              "piecewiseSEM", "sads", "ggplot2", "dplyr", "MASS", "knitr")
instpack <- packages[!packages %in% installed.packages()]
if(length(instpack)>0)
{
    install.packages(packages[!packages %in% installed.packages()])
}
```

One of the packages needed in this tutorial, MCSim (MetaCommunity
Simulation) is not available at CRAN and should be installed directly
from the development repository github. Follow the code bellow to
install it. Information about this package can be found at <a
href="http://rstudio-pubs-static.s3.amazonaws.com/159425_80725873417e42fdb13821c10a198281.html">
MCSim </a>

```{r,  message=FALSE}
if(! "MCSim" %in% installed.packages())
{
    devtools::install_github("sokole/MCSim")
}


```

Attach all packages needed to running this tutorial.


```{r packages, results='hide', message=FALSE}
library(devtools)
library(MCSim)
library(ade4)
library(vegan)
library(lme4)
library(bbmle)
library(optimx)
library(piecewiseSEM)
library(sads)
library(ggplot2)
library(dplyr)
library(MASS)
library(knitr)
source("../code/functions.R")
```

Here we specify the random number generator seed:

```{r seed}
set.seed(42)
```

To evaluate our methods, we generate the meta-community data. Part of
the code used here is adapted from 
[http://rstudio-pubs-static.s3.amazonaws.com/159425_80725873417e42fdb13821c10a198281.html](Sokol's MCSim tutorial).
To replicate our fern data set, we
simulated a metaccomunity of 166 species over in 10 in each of three
regions (mountain chains).

For both deterministic and stochastic scenarios, we simulate 10
different meta-communities until time-step 100 and then take the mean
of the abundances at the 100^ht^ last time-step of each simulation.

```{r generate data}

## Number of sites
Nsites <- 30 
## Number of regions
Nregions <- 3
## Number of species
Nspp <- 153

## Sites attributes ##
## Here are the xy-coordinates of sites: 3 regions with 10 sites each
## Distances between regions is an order of magnitude distance bewteen sites within regions
sites <- expand.grid(x = seq(0,120, length=Nregions), y=seq(1,5, length=Nsites/Nregions))
sites$x <- jitter(sites$x)
sites$y <- jitter(sites$y)
## Each set of 10 points at same x coordinate is labelled as a region
sites$region <- rep(letters[1:Nregions], Nsites/Nregions)
## Enviromental variable: 10 states, that repeat at each site (e.g. altitude)
sites$env <- rep(seq(1,5, length=Nsites/Nregions), each=Nregions)

## Calculate niches for species: optimal values along the enviromental variable
sp.opt <- runif(Nspp, min = 1, max = 5)
```

Here, we generate the sites x species matrix with community initial
conditions to run the simulations.

```{r initial community}
## Initial condition ##
## Initial condition: matrix of sites x spp

## Random values of species abundances 
m0b <- matrix(rlnorm(Nspp*Nsites), Nsites, Nspp)
## Round values of species abundance to represent discrete values of number of individuals
m0b <- round(m0b)
## Splitting species in 3 fractions that are exclusive of each region
R.ind <- sample(letters[1:Nregions], Nspp, replace=TRUE)
for(i in letters[1:Nregions])
  m0b[sites$region==i,R.ind!=i] <- 0
## Calculating Relative abundances
m0b <- sweep(m0b, 1, apply(m0b,1,sum), FUN="/")
```

#### 1.2 Deterministic community

##### 1.2.1 Building the landscape

```{r landscape, results="hide", message=FALSE}
## Following Sokol's tutorial we arbitrarily chose JM = 1e6
JM <- 1e6

## We set m=0.5 to allow some dispersal limitation. Still a half of the deaths are replaced by locals
simulation_landscape_det <- MCSim::fn.make.landscape(
  site.coords = sites[,1:2],
  Ef = sites$env,
  m = 0.5, 
  JM = JM)
## It seems that in R 3.5 it is necessary to convert the dist mat in this object form dataframe to a matrix
simulation_landscape_det$dist.mat <-  as.matrix(simulation_landscape_det$dist.mat)
```
##### 1.2.2 Simulating abundances of deterministic communities

```{r simu det, results="hide", message=FALSE}

## Data frames to store simulations
id <- data.frame(site=rep(1:Nsites, Nspp), sites[,3:4], spp=rep(paste("sp",1:Nspp, sep="."), each=Nsites))
## To store simulation results
Nrep <- 10
det.resu <- matrix(NA, nrow=nrow(id), ncol=Nrep)

for(i in 1:Nrep){
    simu.det <- MCSim::fn.metaSIM( # simulation of deterministic community w correctly observed traits
                           landscape = simulation_landscape_det,
                           ##output.dir.path = 'FERN_SIM_RESULTS_det',
                           ##scenario.ID = 'fern_det',  
                           ##sim.ID = 'det',
                           trait.Ef = sp.opt,
                           trait.Ef.sd = 0.5,
                           J.t0 = m0b,
                           ##gamma.abund = gama.init/(sum(gama.init)),
                           n.timestep = 100, # increased to 100 time steps; initial conditions seems to persist after t=30
                           W.r = 0,
                           nu = 0,
                           speciation.limit = 0,
                           save.sim = FALSE
                       )
    det <- subset(simu.det$J.long, timestep=="100")[,4]
    det.resu[,i] <- det 
}

det <- data.frame(id, count=rowMeans(det.resu))
```




#### 1.3 Stochastic community

##### 1.3.1 Simulating the landscape

  
```{r simu sto, results="hide", message=FALSE}
simulation_landscape_sto <- MCSim::fn.make.landscape(
  site.coords = sites[,1:2],
  Ef = sites$env,
  m = 0.5, 
  JM = JM)
## It seems that in R 3.5 it is necessary to convert
## the dist mat in this object from dataframe to a matrix
simulation_landscape_sto$dist.mat <-  as.matrix(simulation_landscape_sto$dist.mat)
```

##### 1.3.2 Simulating abundances of stochastic communities

```{r sto, results="hide", message=FALSE}
## A matrix for the results
sto.resu <- matrix(NA, nrow=nrow(id), ncol=Nrep)

for(i in 1:Nrep){
    simu.sto <- MCSim::fn.metaSIM(
                           landscape = simulation_landscape_sto,
                           #output.dir.path = 'FERN_SIM_RESULTS_sto_RT',
                           #scenario.ID = 'fern__sto_RT',  
                           #sim.ID = 'sto_RT',
                           trait.Ef = sp.opt,
                           trait.Ef.sd = 1000, # niche deviation changed for neutral dynamics
                           J.t0 = m0b,
                           ##gamma.abund = gama.init/(sum(gama.init)),
                           n.timestep = 100, 
                           W.r = 200, # Dispersal Kernel no longer flat
                           nu = 0,
                           speciation.limit = 0,
                           save.sim = FALSE
                       )
    sto <- subset(simu.sto$J.long, timestep=="100")[,4]
    sto.resu[,i] <- sto 
}

sto <- data.frame(id, count=rowMeans(sto.resu))
```

#### 1.4 Taking samples from simulated communities

Then, we take Poisson samples of the different scenarios. 

Here, we are sampling 50% of simulated communities following Poisson
and negative binomial distributions. Our objective here is to generate
a scenario in which data is sampled from a Poisson distribution,
representing the structure of species abundance data and from a
negative binomial sample, representing the structure of overdispersed
species abundance data.

```{r sampling}
Nsamp <- 0.02
# Poisson samples
## Deterministic community
det.pois <- rpois(length(det$count), det$count*Nsamp)
## Stochastic community
sto.pois <- rpois(length(sto$count), sto$count*Nsamp)

## Looking into results from simulation and sample
par(mfrow=c(2,2))
xtabs(sto$count~sto$spp)%>% as.vector() %>% rad() %>% plot(main="Stochastic simulation")
xtabs(sto.pois~sto$spp)%>% as.vector() %>% rad() %>% plot(main="Stochastic Poisson sample")
xtabs(det$count~det$spp)%>% as.vector() %>% rad() %>% plot(main="Deterministic simulation")
xtabs(det.pois~det$spp)%>% as.vector() %>% rad() %>% plot(rad(det.pois), main="Deterministic Poisson sample")
par(mfrow=c(1,1))
```


### 2. Building hypothesis and models 
<!-- aqui precisa atualizar! --->

We aim to simultaneously test these four main hypotheses:

1. Neutral and niche dynamics affect species abundance

2. Solely neutral dynamics affect species abundance

3. Solely niche dynamics affect species abundance

4. Species abundances vary randomly across regions (null model)

In order to represent each hypothesis, we will build a set of models
using GLMM with fixed and random effects as below. Niche dynamics are
represented by trait and environment effects, whereas neutral dynamics
are represented by species, sites and region effects as random
effects. Given that species traits are measured at species level, we
included species as a random effect on all models (explicitly or
implicity fo models with the term 1+grad|spp). Therefore, this random
effect does not represent neutral or dynamics, but an indiosyncratic
effect of species allowing a random intercept for species abundances.

Hypothesis | Fixed effects | Random Effects
-----------|---------------|----------------
Neutral and niche dynamics | trait + grad + I(grad^2) | 1|spp + 1|spp:site + 1|spp:region + 1+grad|spp 1+grad|site
 | grad + I(grad^2) | 1|spp + 1|spp:site + 1|spp:region + 1+grad|spp + 1+grad|site
Solely neutral dynamics | - | 1|spp + 1|spp:site + 1|spp:region
Solely niche dynamics | trait + grad + I(grad^2)| 1|spp + 1+grad|site
 | grad + I(grad^2)| 1+grad|spp + 1+grad|site
Null model | - | 1|spp + 1|site + 1|region

### 3. Buildind the models 

#### 3.1. Preparing data table for the models

##### Preparing abundance, trait and environmental data

Here, we combine species abundance data sampled from deterministic and
stochastic simulations with traits and environmental data.

```{r selecting comm, results="hide", message=FALSE}

# Binding all data togheter
data <- data.frame(site=id[,"site"], spp=id[,"spp"],
                   det=det.pois, sto=sto.pois)
                   
# Now we need species traits, gradient and spacial info

## Traits 
## A vector with wrong traits with correlation of less than 0.01 with the true traits
cor.t <- 1
while(cor.t>0.01){
    wrong.t <- runif(length(sp.opt), min(sp.opt), max(sp.opt))
    cor.t <- abs(cor(wrong.t, sp.opt))
}
## Trait data
trait.data <- data.frame(spp=unique(id$spp), trait= scale(sp.opt),
                    trait.wr=scale(wrong.t)) # creating vector w/ wrong traits
## Gradient
env.data <- data.frame(site=unique(id$site), grad=scale(sites$env),
                       region=sites$region)

## Preparing data table for model selection
all.data <- merge(data, env.data, by=c("site"))
all.data <- merge(all.data, trait.data, by=c("spp"))
```
A quick check show that all species were included in the sample

```{r check species absence in sample}
samp.tot <- all.data %>%
    group_by(spp) %>%
    summarise(det.tot=sum(det), sto.tot= sum(sto)) 
any(samp.tot$det.tot<=0)
any(samp.tot$sto.tot<=0)
```

Abundances curves of species across the environmental gradient are
as expected:

```{r checking abundance x grad plots}
## A sample of 9 species, a panel for each region
index <- paste("sp",sample(1:nlevels(all.data$spp),9),sep=".")
## Niche
all.data %>%
    filter(spp%in%index) %>%
    ggplot(aes(grad, det)) + geom_line(aes(colour=region)) + facet_wrap(~spp) +
    scale_y_log10() + 
    labs(x="Environmental gradient", y="Species abundance",
         title=paste("Deterministic community from Poisson sample"))
## Neutral
all.data %>%
    filter(spp%in%index) %>%
    ggplot(aes(grad, sto)) + geom_line(aes(colour=region)) + facet_wrap(~spp) +
    scale_y_log10() + 
    labs(x="Environmental gradient", y="Species abundance",
         title=paste("Stochastic community from Poisson sample"))
```


#### 3.2 Adjusting model to the data

First, we create simple functions to perform model selection. Although
the logic is generic, functions below are specific to run this
example. We created a function for each of our hypothesis (see file
'functions.R'). 


In the following sections we fit the models to different sets of data
with these functions. Control parameters were adjusted to avoid
convergence issues.

##### Deterministic community w/ right traits, from poisson sample

```{r models det rt}
## Niche and neutral dynamics
## With traits
nineu.det.rt <- m.full(ab=all.data$det,
                       trait=all.data$trait,
                       grad=all.data$grad,
                       site=all.data$site,
                       region=all.data$region, 
                       spp=all.data$spp,
                       family="poisson",
                       control=glmerControl(optimizer="bobyqa", 
                                            optCtrl=list(maxfun=5e7))
                       )
## Without traits
envneu.det <- m.full2(ab=all.data$det,
                      grad=all.data$grad,
                      site=all.data$site,
                      region=all.data$region, 
                      spp=all.data$spp,
                      family="poisson",
                      control=glmerControl(optimizer="bobyqa", 
                                           optCtrl=list(maxfun=5e7))
                      )
# Niche dynamics
## With traits
niche.det.rt <- m.niche(ab=all.data$det,
                        trait=all.data$trait,
                        grad=all.data$grad,
                        site=all.data$site, 
                        spp=all.data$spp,
                        family="poisson",
                        control=glmerControl(optimizer="bobyqa", 
                                             optCtrl=list(maxfun=5e7))
                        )

## W/o traits
env.det <- m.env(ab=all.data$det,
                 grad=all.data$grad,
                 site=all.data$site,
                 spp=all.data$spp,
                 family="poisson",
                 control=glmerControl(optimizer="bobyqa", 
                                      optCtrl=list(maxfun=5e7))
                 )

## Neutral dynamics
neu.det <- m.neutral(ab=all.data$det,
                     site=all.data$site,
                     region=all.data$region, 
                     spp=all.data$spp,
                     family="poisson",
                       control=glmerControl(optimizer="bobyqa", 
                                            optCtrl=list(maxfun=5e7))
                       )

# Null hypothesis
null.det <- m.null(ab=all.data$det, 
                   site=all.data$site,
                   region=all.data$region, 
                   spp=all.data$spp,
                   family="poisson",
                       control=glmerControl(optimizer="bobyqa", 
                                            optCtrl=list(maxfun=5e7))
                       )


## AIC for each model
AICctab(nineu.det.rt, envneu.det, niche.det.rt, env.det, neu.det, null.det)

```

##### Deterministic community w/ wrong traits, from poisson sample

```{r models det wt}
# Niche and neutral dynamics
nineu.det.wt <- m.full(ab=all.data$det,
                       trait=all.data$trait.wr,
                       grad=all.data$grad,
                       site=all.data$site,
                       region=all.data$region, 
                       spp=all.data$spp,
                       family="poisson",
                       control=glmerControl(optimizer="bobyqa", 
                                            optCtrl=list(maxfun=5e7))
                       )

## Niche dynamics 
niche.det.wt <- m.niche(ab=all.data$det,
                        trait=all.data$trait.wr,
                        grad=all.data$grad,
                        site=all.data$site,
                        spp=all.data$spp,
                        family="poisson",
                        control=glmerControl(optimizer="bobyqa", 
                                             optCtrl=list(maxfun=5e7))
                        )

# AIC for each model
AICctab(nineu.det.wt, envneu.det, niche.det.wt, env.det, neu.det, null.det)
```


##### Stochastic community w/ right traits, from poisson sample

```{r models sto rt}
## Niche and neutral dynamics
nineu.sto.rt <- m.full(ab=all.data$sto,
                       trait=all.data$trait,
                       grad=all.data$grad,
                       site=all.data$site,
                       region=all.data$region, 
                       spp=all.data$spp,
                       family="poisson",
                       control=glmerControl(optimizer="optimx", 
                                            optCtrl=list(method="nlminb",
                                                         starttests = FALSE, kkt = FALSE)))

## Niche and neutral dynamics without species traits
envneu.sto <- m.full2(ab=all.data$sto,
                      grad=all.data$grad,
                      site=all.data$site,
                      region=all.data$region, 
                      spp=all.data$sp,
                      family="poisson",
                      control=glmerControl(optimizer="bobyqa", 
                                           optCtrl=list(maxfun=5e7)))

## Niche dynamics
niche.sto.rt <- m.niche(ab=all.data$sto,
                        trait=all.data$trait,
                        grad=all.data$grad,
                        site=all.data$site,
                        spp=all.data$sp,
                        family="poisson",
                        control=glmerControl(optimizer="bobyqa", 
                                             optCtrl=list(maxfun=5e7)))

## Niche dynamics
env.sto <- m.env(ab=all.data$sto,
                 grad=all.data$grad,
                 site=all.data$site,
                 spp=all.data$sp,
                 family="poisson",
                 control=glmerControl(optimizer="bobyqa", 
                                      optCtrl=list(maxfun=5e7)))

## Neutral dynamics
neu.sto <- m.neutral(ab=all.data$sto,
                     site=all.data$site,
                     region=all.data$region, 
                     spp=all.data$sp,
                     family="poisson")
## Null hypothesis
null.sto <- m.null(ab=all.data$sto, 
                   site=all.data$site,
                   region=all.data$region, 
                   spp=all.data$sp,
                   family="poisson")

# AIC for each model
AICctab(nineu.sto.rt, envneu.sto, niche.sto.rt, env.sto, neu.sto, null.sto)

```



### 4. Addressing main criticisms of our framework

We now will compare results from model selection of the different
scenarios we created. We consider the model with lowest AIC (Akaike
Information Criterion) as the model that best explains the community
data. We will depict all best models in terms of their adjusted
$R^{2}$. Adjusted $R^{2}$, as they measure the relative importance of
fixed and random effects in the model, are used here a proxy of
correspondence of community processes to terms in the
model. Therefore, conditional $R^{2}$ represents the influence of
niche dynamics whereas marginal $R^{2}$ is partitioned into each
random term and can represent either niche (in the case of the term
1+grad|spp and 1+grad|site), neutral dynamics (for 1|spp:mountain and
1|spp:site) or idiosyncratic effects of species (1|spp).

We adapted scripts from Johnson (2014) to calculate marginal and
conditional $R^{2}$ of each models, and also the enhanced agreement
repeatibility (Stofel et al 2107), or the ratio of the intra-class
variance for a given random factor and the total variance estimated by
the models, including fixed-effect variances. These functions are
available in the script 'functions.R', but applies only to the models
fitted in this tutorial. For generic functions please see packages
(rptR and MuMIn).


#### 4.1. Are fixed and random effects actually capturing niche and neutral dynamics?

In order to address this question, we compare results of the selected
model for deterministic and stochastic communities with traits
strongly correlated with species abundance sampled from Poisson
distribution.

```{r test 4.1}
#Calculating R2
r2.det.rt <- r2.full(nineu.det.rt, null.model = neu.det)$full.table
r2.sto <- r2.neutral(neu.sto, null.model = neu.sto)$full.table

table1 <- merge(r2.det.rt, r2.sto, by=c("component", "type"), all.x=TRUE, all.y=TRUE)
colnames(table1) <- c("Effect", "Process", "R2 Deterministic community w/ right traits", 
                   "R2 Stochastic community w/ right traits")
table1 <- table1[order(table1$Process),]
```


```{r table1, echo=FALSE, results='asis'}
kable(table1)
```

For communities built by known niche dynamics conditional $R^{2}$
value is composed basically by fixed effects, representing niche
dynamics. For communities built by known neutral dynamics, conditional
$R^{2}$ is built by random effects, specially the effect represented
by the term (1|region:spp), which express limited dispersal of
species within regions.

#### 4.2. Are random effects only capturing neutral dynamics or uninformed species traits are inflating random effects? 

In order to examine how random effects reflect uninformed traits, we
compare the results of the selected model from deterministic
communities with right and wrong traits.

```{r test 4.2 right-wrong}
r2.det.wt <- r2.env(env.det, null.model = neu.det)$full.table

## Creating table
table2 <- merge(r2.det.rt, r2.det.wt, by=c("component", "type"), all.x=TRUE, all.y=TRUE)
colnames(table2) <- c("Effect", "Process", "R2 Deterministic community w/ right traits", 
                   "R2 Deterministic community w/ wrong traits")
table2 <- table2[order(table2$Process),]
```

```{r table2, echo=FALSE, results='asis'}
kable(table2)
## Most of conditional R2 goes to 1|spp and not to 1|spp:site or 1|spp:region, check
```

Only looking into the $R^{2}$ conditional values, one cannot examine
difference between deterministic communities built by uninformative
and informative species traits. However, by examining the partition
$R^{2}$ values, one can detect if traits used in the model are
actually influencing species abundances. Uninformative traits inflate
marginal $R^{2}$ because of the importance of the term related to the
random effects of species (env|spp).  The influence of the
trait-environment interaction ('trait:grad') declines, and so the
repeatability of fixed effects, that include this interaction. When
traits are informative and niche dynamics are preponderant, $R^{2}$
for fixed effects should be particularly high.




